---
title: "All about aRboretum"
subtitle: "aRboretum: Complete technical guide for botanical collections"
author: "Martin Boucknooghe & Domingos Cardoso"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aRboretum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

`aRboretum` is an integrated R package designed to automate the production of botanical collection labels, from taxonomic data extraction to interactive web labels and print-ready QR codes. It can be used for arboreta, botanic gardens, educational trails, and other living plant collections.

The package offers:

- multilingual output in Portuguese, English, French, and Spanish;
- synonym resolution and integration of data from [Flora e Funga do Brasil (FFB)](https://floradobrasil.jbrj.gov.br/) and the [World Checklist of Vascular Plants (WCVP)](https://powo.science.kew.org/), the backbone of Plants of the World Online (POWO);
- global conservation status from the [IUCN Red List](https://www.iucnredlist.org/);
- browser-based guides to review the retrieved data and edit the species phrases;
- interactive HTML labels with species information, text-to-speech, distribution maps, photo slideshows, and optional personal audio recordings;
- a searchable minisite index linking all generated labels;
- printable QR-code labels in different layouts;
- one flexible additional language slot for community-based accessibility through `add_lang`.

All functions are designed to work together. This vignette provides a technical overview of the package workflow, main arguments, expected data structure, and common customisation options.


## Installation

`aRboretum` depends on two packages that are not on CRAN, so install them first:

- [`rWCVPdata`](https://github.com/matildabrown/rWCVPdata), which provides the World Checklist of Vascular Plants data and is distributed through its own repository;
- [`floraR`](https://github.com/DBOSlab/floraR), which downloads and parses the Flora e Funga do Brasil dataset used by `arboretum_data()`.

```{r}
if (!requireNamespace("rWCVPdata", quietly = TRUE)) {
  install.packages("rWCVPdata",
                   repos = c(
                     "https://matildabrown.github.io/drat",
                     "https://cloud.r-project.org"
                   )
  )
}

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Install the development version of floraR from GitHub,
# together with its required dependencies
BiocManager::install("DBOSlab/floraR", dependencies = TRUE)
```

Then install the development version of `aRboretum` from [GitHub](https://github.com/DBOSlab/aRboretum):

```{r}
# Install the development version of aRboretum from GitHub,
# together with its required dependencies
BiocManager::install("DBOSlab/aRboretum", dependencies = TRUE)
```

```{r setup}
library(aRboretum)
```

### Optional: IUCN Red List API token

By default, IUCN Red List categories are retrieved from the IUCN Red List checklist mirrored by [GBIF](https://www.gbif.org/), which requires no registration. To query the official [IUCN Red List API](https://api.iucnredlist.org/) instead, install the [`rredlist`](https://docs.ropensci.org/rredlist/) package, request a free API token, and store it in your `.Renviron` file:

```{r}
install.packages("rredlist")
usethis::edit_r_environ()   # add the line: IUCN_REDLIST_KEY=your_token
```


## Internal data structures

#### The internal translation dictionary

The package contains an internal translation table used to build species descriptions in the built-in languages. It includes terms and sentence fragments related to:

- taxonomy;
- endemism and establishment means;
- phytogeographic domains and vegetation types;
- IUCN categories;
- genus diversity notes;
- sentence-building connectors.

The helper functions `.tr_dict()` and `.phrase_generator()` use this dictionary to produce natural-language species summaries.

#### Botanical regions mapping

The package also relies on an internal mapping between botanical regions (the level-3 areas of the World Geographical Scheme for Recording Plant Distributions used by WCVP), countries, and continents, with their names in the four built-in languages. This allows WCVP distribution data to be converted into readable geographic statements and distribution maps.


## Expected data structure

Most functions rely on the standard species data file written by `arboretum_data()` and completed by `arboretum_phrases()`. Its columns are:

```text
# Written by arboretum_data()
family
genus
taxonName
scientificNameAuthorship
FFB.vernacularName
country
endemism                  # endemism in Brazil
botanical_country
introduced_to
FFB.establishmentMeans
FFB.stateProvince
FFB.phytogeographicDomain
FFB.vegetationType
FFB.genusRichness
FFB.genusRank
IUCN.status
plant_uses_EN, plant_uses_PT, plant_uses_ES, plant_uses_FR
free_notes_EN, free_notes_PT, free_notes_ES, free_notes_FR
POWO.url
FFB.url

# Added by arboretum_phrases()
full_phrases_EN, full_phrases_PT, full_phrases_ES, full_phrases_FR
full_phrases_ADD_LANGUAGE
```

Not every column needs to be filled, but the core taxonomic and distribution fields are required for full functionality. `arboretum_audios()` and `arboretum_labels()` also require the `full_phrases_*` columns of every language they use.


## Main workflow

A typical `aRboretum` workflow consists of the following steps:

1. extract and structure species data with `arboretum_data()`, then review it;
2. generate the editable species phrases with `arboretum_phrases()`;
3. optionally prepare folders for personal audio with `arboretum_audios()`;
4. optionally prepare folders for species photos with `arboretum_photos()`;
5. generate interactive HTML labels with `arboretum_labels()`;
6. create a searchable minisite with `arboretum_minisite()`;
7. create printable QR-code labels with `arboretum_qrcodes()`.


## Function overview

## `arboretum_data()`

### What it does

`arboretum_data()` extracts and merges species information from Flora e Funga do Brasil and the World Checklist of Vascular Plants, and adds the IUCN Red List status of each species. It resolves synonyms when possible, merges overlapping information, and returns a standardised dataset that can be used by the rest of the package.

The output also includes empty user-editable columns such as `plant_uses_*` and `free_notes_*`, which can later be used to enrich labels in the built-in languages.

### Example

```{r}
spp_list <- c("Luetzelburgia bahiensis", "Paubrasilia echinata")

extracted_data <- arboretum_data(
  spp_list = spp_list,
  printed_lang = c("pt", "en", "fr", "es"),
  save = TRUE,
  format = "xlsx",
  dir = "arboretum_data",
  ffb_dir = "flora_download"
)
```

### Data reviewing guide

When `save = TRUE`, the function also writes `__data_reviewing_guide.html` to `dir`. It shows one card per species with editable fields for the vernacular names, country, endemism in Brazil, botanical countries, introduced range, establishment means, Brazilian states, phytogeographic domains, vegetation types, IUCN status, plant uses, and free notes. Endemism, establishment means, and IUCN status are drop-down menus restricted to the values used to build the phrases. The corrected data can be saved or exported as CSV or XLSX directly from the browser. `printed_lang` sets the languages of the guide interface and which plant-use and note fields it shows; the data file always keeps the columns of all four languages.

### Notes

- Synonyms are resolved to accepted names when possible.
- Missing values from one source may still be complemented by the other.
- The FFB dataset is downloaded with `floraR` into `ffb_dir` and kept there. Later runs always use the most recent version already downloaded, without contacting the FFB server, so the function also works offline. To update it, run `floraR::flora_download(dir = ffb_dir)`. `ffb_dir` can also point to one specific version folder, such as `"flora_download/dwca_ffb_v393_430_latest"`.
- WCVP data come from the `rWCVPdata` package, so no per-species web request is needed.
- If `dir` already contains a data file, it is read instead of querying the databases again, and the reviewing guide is rebuilt from it.
- No phrase is generated at this step.

## `arboretum_phrases()`

### What it does

`arboretum_phrases()` generates the natural-language species descriptions once the data retrieved with `arboretum_data()` have been reviewed. The phrases are stored in the data file itself, in the columns `full_phrases_EN`, `full_phrases_PT`, `full_phrases_ES`, `full_phrases_FR`, and `full_phrases_ADD_LANGUAGE`, which are created at this step only. It also writes an HTML phrase guide (`__phrase_generating_guide.html`) next to the data file, in which phrases, vernacular names, plant uses and notes can be edited and exported.

### Example

```{r}
arboretum_phrases(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en", "fr", "es"),
  add_lang = "TUKANO"
)
```

### Notes

- Stored phrases, including manual edits, are never overwritten; only empty phrases are filled when the function is run again.
- `overwrite = TRUE` regenerates all phrases, for example after correcting the data. `full_phrases_ADD_LANGUAGE` is never overwritten.
- Phrases pick some vernacular names and countries at random, so regenerating a phrase may change its wording; storing them keeps them stable.
- `arboretum_audios()` and `arboretum_labels()` only read the stored phrases and never generate them.

## `arboretum_audios()`

### What it does

`arboretum_audios()` prepares a folder structure for personal species audio recordings. For each species and each language, the function creates one folder, inside `arboretum_audios/`, where users can place audio files.

It also writes an HTML recording guide, `arboretum_audios/__personal_audio_recording_guide.html`, showing the phrase stored by `arboretum_phrases()` for each species and language. In Chromium-based browsers, the guide can also record directly into the species folders.

### Example

```{r}
arboretum_audios(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en", "fr", "es"),
  add_lang = "TUKANO",
  verbose = TRUE
)
```

### Notes

- Built-in languages are `"pt"`, `"en"`, `"fr"`, and `"es"`.
- The optional `add_lang` argument creates one additional language folder per species, such as `TUKANO`.
- Species folders are only created when `arboretum_audios/` does not exist yet; the recording guide is rewritten on every run.

## `arboretum_photos()`

### What it does

`arboretum_photos()` creates one photo folder per species so that users can add personal or curated images to be embedded later in the HTML labels.

### Example

```{r}
arboretum_photos(
  data_path = "arboretum_data/arboretum_data.xlsx",
  verbose = TRUE
)
```

### Notes

- Existing folders are preserved, and only missing species folders are created.
- Photos are optional and only appear in labels when files are found in the expected species folder.

## `arboretum_labels()`

### What it does

`arboretum_labels()` generates one standalone HTML label per species. Each label may include:

- multilingual species text, taken from the phrases stored by `arboretum_phrases()`;
- plant uses and free notes;
- taxonomic information;
- world and Brazil distribution maps;
- browser-based speech synthesis;
- personal audio recordings when available;
- photo slideshows when available;
- IUCN iconography and links to FFB and POWO;
- institutional and package logos.

### Example

```{r}
arboretum_labels(
  data_path = "arboretum_data/arboretum_data.xlsx",
  audio_dir = "arboretum_audios",
  photo_dir = "arboretum_photos",
  printed_lang = c("pt", "en"),
  add_lang = "TUKANO",
  path_to_logo = "path/to/institution_logo.png",
  logo_url = "https://www.example.org",
  dir = "species_labels",
  verbose = TRUE
)
```

### The `add_lang` workflow

The `add_lang` argument provides a flexible, interim pathway for including one additional community or local language in species labels and associated audio, without requiring translation of the full website interface.

When `add_lang` is supplied:

- `arboretum_phrases()` adds an editable field for the extra language to the phrase guide, stored in the column `full_phrases_ADD_LANGUAGE`;
- `arboretum_labels()` looks for the column `full_phrases_ADD_LANGUAGE`;
- if that column contains non-empty text, the extra language is added as an additional text option in the label;
- the custom language text is read directly from `full_phrases_ADD_LANGUAGE`;
- this extra language does **not** pass through the standard phrase-generation and translation workflow.

This is especially useful in collaborative work with Indigenous peoples or other local communities, where species-level accessibility may be more important than translating the full minisite interface.

### Notes

- The function stops if the `full_phrases_*` column of a language in `printed_lang` is missing, and warns about species without a stored phrase.
- Personal audio files take priority over browser speech synthesis.
- For custom languages added through `add_lang`, personal recordings are recommended because browser voices may not be available.
- Personal audio and photo folders are copied into `dir` (as `__arboretum_audios/` and `__arboretum_photos/`), and photos larger than 1 MB are compressed in the copy.
- Photos and maps are included only when available.

## `arboretum_minisite()`

### What it does

`arboretum_minisite()` builds a searchable `index.html` page linking all generated species labels. When the species data file is provided, it also includes a dashboard with collection summaries: numbers of species, genera and families, establishment means, phytogeographic domains, Brazilian states, and countries.

### Example

```{r}
arboretum_minisite(
  labels_dir = "species_labels",
  data_path = "arboretum_data/arboretum_data.xlsx",
  site_title = "Garden Collection",
  printed_lang = c("en", "pt", "es", "fr"),
  logo = "path/to/logo.png",
  logo_url = "https://www.example.org",
  group_by_family = TRUE
)
```

### Notes

- The interface language order follows the order of `printed_lang`.
- The minisite can work without `data_path`, but vernacular names and the dashboard are then omitted.
- After writing `index.html`, the labels folder is renamed after the site title, here `Garden_Collection_minisite/`. The whole folder can then be published as a website.

## `arboretum_qrcodes()`

### What it does

`arboretum_qrcodes()` creates printable A4 sheets of QR-code labels that can link to the species HTML labels of a published minisite, to external URLs, or to source records. Four layouts are available: `"minimalist"` (QR code with the species and vernacular names) and the detailed `"classic"`, `"modern"`, and `"botanical"` layouts (with authorship, family, endemism, optional logo and specimen code).

### Example

```{r}
arboretum_qrcodes(
  data_path = "arboretum_data/arboretum_data.xlsx",
  base_url = "https://myinstitution.org/collection",
  layout = "minimalist",
  dir = "qrcodes_mini"
)

arboretum_qrcodes(
  data_path = "arboretum_data/arboretum_data.xlsx",
  base_url = "https://myinstitution.org/collection",
  layout = "classic",
  width = 6.0,
  length = 8.0,
  id_code = "catalogue_number",
  printed_lang = "en",
  path_to_logo = "path/to/logo.png",
  color = "#1a2e1a",
  format = "pdf",
  dir = "qrcodes_classic"
)
```

### Notes

- QR codes point, in order of priority, to `base_url` (one page per species, matching the label file names), to `url`, to the POWO page, to the FFB page, or encode the taxon name.
- `id_code` can be the name of a column of the data file or one or more codes.
- Detailed layouts print a single language, the first one in `printed_lang`.
- Default label sizes are 2.5 × 2.5 cm for the minimalist layout and 6 × 8 cm for the detailed layouts.
- Output can span multiple pages automatically, as PDF or PNG.

## Practical example with one additional language

The example below illustrates a simple workflow where Portuguese and English remain the main interface languages, while Tukano is added as a community language for species labels and optional personal recordings.

```{r}
arboretum_phrases(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en"),
  add_lang = "TUKANO"
)

arboretum_audios(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en"),
  add_lang = "TUKANO"
)

arboretum_labels(
  data_path = "arboretum_data/arboretum_data.xlsx",
  audio_dir = "arboretum_audios",
  printed_lang = c("pt", "en"),
  add_lang = "TUKANO",
  dir = "species_labels"
)
```

Before generating the labels, fill the column `full_phrases_ADD_LANGUAGE` created by `arboretum_phrases()` with the full text to display in Tukano for each species, either in the phrase guide or directly in the data file.


## Defensive checks and internal helpers

Several internal helper functions provide validation and error prevention.

### `.arg_check_spp_list()`

- trims whitespace;
- standardises input species names;
- checks for invalid entries.

### `.arg_check_dir()`

- validates output directory names;
- removes trailing slashes when needed.

### `.arg_check_printed_lang()`

- standardises language codes to lower case;
- removes duplicates;
- validates built-in language codes.

### `.read_species_data()`

- checks file existence and extension;
- reads CSV or Excel input;
- validates required columns.

### `.load_ffb()`, `.load_wcvp()`, and `.get_iucn_status()`

- load the FFB dataset, reusing a previous download;
- load the WCVP tables once per run;
- retrieve IUCN Red List categories from GBIF or the IUCN Red List API.

### `.copy_folder_progress()`

- copies user-provided audio or photo folders into output directories;
- shows progress feedback when many files are involved.

### Phrase-generation helpers

Helpers such as `.phrase_generator()`, `.tr_dict()`, `.botdiv_to_countries()`, and `.add_genus_curiosity_notes()` support the construction of species descriptions in the built-in languages.

## Conclusion

`aRboretum` provides a practical pipeline for creating modern botanical collection labels, from taxonomic data retrieval to multilingual HTML pages and printable QR codes. The package combines automation with flexibility, allowing users to review the data and edit the phrases, and to enrich labels with notes, uses, photos, personal audio, and one additional community-based language through `add_lang`.

For function-specific details, see the package documentation:

```r
?arboretum_data
?arboretum_phrases
?arboretum_audios
?arboretum_photos
?arboretum_labels
?arboretum_minisite
?arboretum_qrcodes
```
