---
title: "All about aRboretum"
subtitle: "aRboretum: Complete technical guide for botanical collections"
author: "Martin Boucknooghe & Domingos Cardoso"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aRboretum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

aRboretum is an integrated R package designed to automate the production of botanical collection labels, from data mining to interactive web labels and print‑ready QR codes. It targets herbarium, arboretum, and all living plant collections, offering:

- Multilingual output (Portuguese, English, French, Spanish)

- Synonym resolution and data merging from two sources (Flora e Funga do Brasil and Plants of the World Online)

- Interactive HTML labels with species information, text‑to‑speech, distribution maps (global + Brazilian states), photo slideshows, and personal audio fallback

- Searchable minisite index with filtering and summary dashboard

- Print‑ready QR code labels in two layouts (minimalist/complete) with custom branding, specimen IDs, and automatic tiling on A4 pages

All functions are designed to work together seamlessly. This vignette provides a complete overview, including data handling, error prevention, and customisation.

### Installation & required packages

```{r setup}
# Install from GitHub
if (!require("devtools")) install.packages("devtools")
devtools::install_github("DBOSlab/aRboretum")
```

R package dependencies (automatically installed):
floraR, taxize, openxlsx, dplyr, tidyr, sf, geobr, leaflet, htmltools, base64enc, jsonlite, qrcode, png, jpeg, grid, grDevices, stringi, tools, magrittr.

## Internal data structures

#### The dict translation table
The package contains an internal dict tibble that maps keys to phrases or words in four languages. It covers:

- Taxonomic statements (“belongs to”, “family”)
- Distribution and endemism (“endemic to”, “non‑endemic”)
- Establishment means (“native”, “cultivated”, “naturalised”)
- Phytogeographic domains (“Amazon”, “Atlantic Forest”, etc.)
- Vegetation types (carrasco, restinga, igapó, etc.)
- IUCN status labels (EN, VU, etc.)
- Genus diversity informations
- Sentence-building words

The auxiliary function .tr_dict(key, lang, dict) retrieves the correct translation. .phrase_generator() uses this dictionary to build natural‑language species descriptions.

#### Botanical regions mapping

The package also includes an internal dataset botregions (not shown in the code but referenced in .phrase_generator and .botdiv_to_countries). It maps:

- botanical divisions → countries → continents
- Translations for botanical division names (e.g., “BRA” → “Brazil” in English, “Brasil” in Portuguese), countries and continents

This allows converting POWO’s three‑letter codes into human‑readable country names and generating accurate global distribution statements.

#### Data columns expected

All functions rely on a standard data frame produced by arboretum_data_miner(). Required columns include:

```text
family, taxonName, scientificNameAuthorship, FFB.vernacularName, country, endemism, FFB.establishmentMeans, FFB.stateProvince, FFB.phytogeographicDomain, FFB.vegetationType, botanical_country, introduced_to, IUCN.status, plant_uses_EN, plant_uses_PT, plant_uses_ES, plant_uses_FR, free_notes_EN, free_notes_PT, free_notes_ES, free_notes_FR, POWO.url, FFB.url
```
Missing columns trigger an explicit error.

### Details of aRboretum's functions

```text
arboretum_data_miner()
```
#### What it does :
- Downloads the latest Flora e Funga do Brasil Darwin Core Archive (DwC‑A), a large file cached temporarily.

- Parses four tables: taxon.txt, distribution.txt, vernacularname.txt, speciesprofile.txt.

- For each species in spp_list:
  - Searches for exact name match in FFB.
  - If the name is a synonym (taxonomicStatus == "SINONIMO"), retrieves the accepted name ID and uses that for all subsequent data.
  - Merges data from the four tables.
  - Also queries POWO via taxize::pow_search() and pow_lookup(), resolves synonyms there too.

- Merges FFB and POWO data, giving priority to FFB for overlapping fields (e.g., vernacular names, Brazilian states).

- Adds genus diversity metrics (number of species per genus in Brazil, rank).

- Add two new types of empty columns named “free_notes” and “plant_uses” each containing four columns for each language supported by aRboretum (e.g., free_notes_EN, plant_uses_ES). These fields are intended to be filled in manually by users. 

- Saves a CSV or Excel file (if save = TRUE).

- Cleans up temporary flora_download folder.


#### Potential problems handled :

- Missing species in one database: fields become NA; the other database fills what it can.
- Synonyms: the function automatically resolves to the accepted name.
- Empty strings are converted to NA for consistency.
- Duplicated accepted names in POWO results are deduplicated.
- Large download: only one DwC‑A download per function call; folder deleted after processing each species to avoid disk bloat.
- Empty FFB results: the merged result still works with POWO‑only data.

#### Example :
```{r}
species_list <- c("Luetzelburgia bahiensis", "Paubrasilia echinata")

mined_data <- arboretum_data_miner(
  spp_list = spp_list,
  save = TRUE,
  format = "csv",
  dir = "arboretum_mined_data"
)
```

#### Options: 

Two empty columns named “free_notes” and “plant_uses” have now been added to the dataframe. You can use these columns to add any information you find relevant. 
Translations are entered manually into the columns for the different languages supported by the package (en, pt, fr, es). 
The added sentences will later be placed within the species records in sections defined in the “Overview of the species” section.


```text
arboretum_personal_audios()
```
#### What it does :

- Reads the mined data.

- Creates a folder arboretum_personal_audios.

- For each species and each language in printed_lang, creates a subfolder named FAMILY_Genus_species_LANG (e.g., FABACEAE_Paubrasilia_echinata_EN).

- Generates a single text file __species_phrases.txt containing the full descriptive phrase (without HTML tags) for every species in every language – a ready‑to‑read script for recording.

#### Potential problems handled :

- Existing directories: if the base folder already exists, the function does not recreate it (preserves user‑added audio files). Only missing species/language folders are created.
- Empty data: skips species with missing required fields.
- Encoded text: the phrase file is written in UTF‑8 to support special characters.

#### Example :

```{r}
arboretum_personal_audios(
  data_path = "arboretum_mined_data/arboretum_mined_data.csv",
  printed_lang = c("pt", "en", "fr", "es"),   # default all four
  verbose = TRUE
)
```


```text
arboretum_photos()
```
#### What it does :

- Reads the mined data.

- Creates a folder arboretum_photos.

- For each species, creates a subfolder named FAMILY_Genus_species_photos, where you can upload personal or external photos of the relevant species.

#### Potential problems handled :

- Existing folders: does not overwrite; creates only missing ones.
- Invalid characters: the species names  replaced by underscores).

#### Example :

```{r}
arboretum_photos(
  data_path = "arboretum_mined_data/arboretum_mined_data.xlsx",
  verbose = TRUE
)
```


```text
arboretum_labels()
```
#### What it does :

- Loads the mined data.

- Generates HTML phrases for each language using the auxiliary function .phrase_generator(), which builds natural language descriptions from the data and the dictionary translations.

- For each species:
  - Builds distribution maps (world map using WGSRPD level 3 & Brazil state map using geobr data).
  - Checks for personal audio files (by scanning the folder previously created by arboretum_personal_audios). If found, embeds them as Base64 data URIs, otherwise uses Web Speech API.
- Checks for personal photos (scans arboretum_photos subfolder). If found, creates a Ken‑Burns slideshow.
- Embeds logos (institutional logo if provided, plus package logos for JBRJ, MMA, Reflora, aRboretum).
- Adds IUCN status icon (from package figures).
- Writes a standalone HTML file with CSS, Javascript for language switching, speech synthesis, and slideshow controls.

#### Potential problems handled :

- Missing maps: If no state occurrence data, only the world map is shown. If world map download fails, the label still works (no map).
- Personal audio missing: falls back to synthesized voices.
- Personal photos missing: no slideshow section.
- Unsupported audio formats: only recognised extensions (mp3, wav, m4a, ogg, flac) are used.
- Voice availability: the script prioritise the personnal voices for each language; falls back to generic if exact match missing.

Note: audio and photos are Base64‑encoded, which can increase HTML file size, acceptable for moderate numbers.


#### Example :

```{r}
arboretum_labels(
  data_path = "arboretum_mined_data/arboretum_mined_data.xlsx",
  personal_audio_dir = "arboretum_personal_audios",
  photo_dir = "arboretum_photos",
  printed_lang = c("pt", "en"),
  path_to_logo = "path/to/institution_logo.png",
  dir = "species_labels"
)
```


```text
arboretum_minisite()
```
#### What it does :

- Scans a directory of HTML label files (produced by arboretum_labels).

- Parses filenames to extract family and species (pattern: FAMILY_Genus_species_label.html).

- If data_path is supplied, joins vernacular names and computes dashboard statistics (total species, genera, families, establishment counts, state counts, phytogeographic domain counts, country counts).

- Builds a single index.html page with:
  - Real‑time search (species, family, vernacular).
  - Language toggle buttons (re‑using the same dictionary), the order of the language buttons depends on the order of the arguments (e.g., c(“pt”, “en”); the default language will be Portuguese) .
  - Dashboard charts (donut for origins & domains, bar charts for states & countries).
  - Responsive grid of species cards, optionally grouped by family.
  - Logo Base64: embedded so that the index is fully self‑contained, may contain a link to a source site.
  - Footer with aRboretum credit.

#### Potential problems handled :

- Missing data_path: dashboard is omitted, but the species list still works.
- No label files: stops with a clear error.
- Common names separated by vertical bars: all common names are displayed.
- Filename parsing: robust to trinomials (e.g., ORCHIDACEAE_Cattleya_trianae_var_alba_label.html).
- Renaming of labels directory: the function renames the original labels_dir to a sanitised version of site_title (spaces replaced by underscores) to avoid conflicts when deploying.

#### Example :

```{r}
arboretum_minisite(
  labels_dir = "species_labels",
  data_path = "arboretum_mined_data/arboretum_mined_data.xlsx",
  site_title = "Personnal Garden Collection",
  printed_lang = c("en", "pt", "es", "fr"),
  logo       = "path/to/logo.png",
  logo_url   = "https://www.my.logos.website",
  group_by_family = TRUE
)
```


```text
arboretum_qrcodes()
```

####The aim:

This feature allows us to link species records to their own unique QR code. To do this, you need to create your own URL and set up a website that the QR code will link to. The microsite created earlier will therefore serve as the main interface for your site.

#### What it does :

- Reads the mined data (optional species filter).

- For each species, determines the URL to encode in the QR code:

  1. If base_url is provided, builds a link to the corresponding
  label file (base_url/FAMILY_Genus_species_label.html).
  2. Else if url is provided, uses that (recycled if necessary).
  3. Else uses POWO.url column.
  4. Else uses FFB.url column.
  5. Else uses the taxon name as plain text.

- Computes A4 page layout (margins, gaps) based on label dimensions (customisable via width and length).

- Generates QR matrices using qrcode::qr_code().

- Converts QR matrix to a coloured raster.

- Draws either minimalist or complete layout using grid graphics.
  - Minimalist: species name (italic), first vernacular name (small, grey), thin separator, QR code.
  - Complete: species name + authorship, family (bold, small caps), first vernacular name, endemism phrase (translated according to printed_lang), QR code, optional specimen ID (bottom‑left), optional institutional logo (bottom‑right).

- Tiles labels on A4 pages, saves as PDF or PNG.

#### Potential problems handled :

- Missing URLs: gracefully falls back to taxon name.
- Logo loading: supports PNG and JPEG, if file missing or unsupported, logo omitted (with warning).
- Aspect ratio preservation: logo scaled to fit within allocated space.
- Label size adaptation: font sizes and QR side length are dynamically scaled based on width and height.
- Quiet zone: two white modules added around QR code to ensure scannability.
- ID codes: can be taken from a column name or a vector recycled to number of species.
- Multi‑page output: automatically creates pageXX files when species exceed per‑page capacity.

#### Example :

```{r}
# Minimalist (default 2.5x2.5 cm)
arboretum_qrcodes(
  data_path = "arboretum_mined_data/arboretum_mined_data.xlsx",
  base_url = "https://myinstitution.org/collection",
  layout = "minimalist",
  dir = "qrcodes_mini"
)

# Complete with specimen IDs
arboretum_qrcodes(
  data_path = "arboretum_mined_data/arboretum_mined_data.xlsx",
  base_url = "https://myinstitution.org/collection",
  layout = "complete",
  width = 6.0, length = 8.0,
  id_code = "catalogue_number",   # column in the data
  printed_lang = "en",
  path_to_logo = "path/to/logo.png",
  color = "#1a2e1a",
  format = "pdf",
  dir = "qrcodes_complete"
)
```

Internal auxiliary defensive mechanisms
The auxiliary functions provide many safety checks:

```text
.arg_check_spp_list:
```
- Trims whitespace, capitalises first letter of each string.
- Ensures every element contains a space.
- Stops with a helpful error message listing invalid entries.

```text
.arg_check_dir:
```
- Verifies input is a single character string, not empty, removes trailing slash.

```text
.arg_check_printed_lang:
```
- Converts to lowercase the lang arguments, removes duplicates, checks against allowed values c("pt","en","fr","es").
- Stops if invalid codes are present.

```text
.read_species_data:
```
- Checks file existence and extension (csv/xlsx).
- Validates presence of all required columns.
- Returns a meaningful error if reading fails.

```text
.get_accepted_name_id:
```
- Queries taxonomic status; if synonym, retrieves accepted name ID and prints a message (if verbose).

```text
.botdiv_to_countries:
```
- Handles truncation of long botanical division names (POWO sometimes shortens).
- Merges multiple countries per division with " | ".

```text
.add_genus_curiosity_notes:
```
- Adds notes only if you choose to populate the “free notes” column in your dataframe or if relevant information about the genus is available
- Uses translated expressions from the dictionary for mentions related to genus diversity.

```text
.tr_dict and .tr_dict_vec:
```
- Uses the `dict` dictionary to translate the various words contained in the species sentences based on keys associated with them.
- Returns character strings; if the key is missing, returns NA (which should not occur with a complete dictionary).

```text
.convert_acronym_br_state:
```
- Converts state acronyms (e.g., "SP") to full names using Unicode‑safe matching (normalises diacritics).
- Handles both acronym and full name inputs.

```text
.copy_folder_progress:
```
- Shows progress bar when copying many files (e.g., personal audio directory).
- Creates destination folder if missing.

QR code drawing helpers:

- ```text .qr_to_raster() ``` adds a quiet zone of 2 modules.
- ```text .load_logo_raster() ``` tries to load PNG/JPEG, warns on failure.
- Both layouts include extensive viewport calculations to avoid clipping.


## Conclusion

The `aRboretum` package provides a robust, production‑ready pipeline that handles real‑world complexities: large data downloads, synonym resolution, multilingual output, personal media integration, and print layout. Every function includes defensive checks to avoid common mistakes and gives informative error messages. By following this vignette, you can fully automate the creation of modern, accessible botanical labels for your collection.

For further details, see the function‑level documentation (`?arboretum_data_miner`, etc.). For bug reports or feature requests, please use the GitHub issue tracker.
