---
title: "Quick start with aRboretum"
subtitle: "How to generate interactive species HTML labels and a minisite"
author: "Martin Boucknooghe & Domingos Cardoso"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```


## Introduction

This vignette provides a concise overview of the core `aRboretum` workflow: retrieving and reviewing species data, generating the species phrases, building interactive HTML labels, and creating a searchable minisite for a living plant collection.

### Installation

`aRboretum` depends on two packages that are not on CRAN, so install them first:

- [`rWCVPdata`](https://github.com/matildabrown/rWCVPdata), which provides the World Checklist of Vascular Plants data and is distributed through its own repository;
- [`floraR`](https://github.com/DBOSlab/floraR), which downloads and parses the Flora e Funga do Brasil dataset used by `arboretum_data()`.

```{r}
if (!requireNamespace("rWCVPdata", quietly = TRUE)) {
  install.packages("rWCVPdata",
                   repos = c(
                     "https://matildabrown.github.io/drat",
                     "https://cloud.r-project.org"
                   )
  )
}

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Install the development version of floraR from GitHub,
# together with its required dependencies
BiocManager::install("DBOSlab/floraR", dependencies = TRUE)
```

Then install the development version of `aRboretum` from [GitHub](https://github.com/DBOSlab/aRboretum):

```{r}
# Install the development version of aRboretum from GitHub,
# together with its required dependencies
BiocManager::install("DBOSlab/aRboretum", dependencies = TRUE)
```

#### Optional: IUCN Red List API token

By default, IUCN Red List categories are retrieved from the IUCN Red List checklist mirrored by [GBIF](https://www.gbif.org/), which requires no registration. To query the official [IUCN Red List API](https://api.iucnredlist.org/) instead, install the [`rredlist`](https://docs.ropensci.org/rredlist/) package, request a free API token, and store it in your `.Renviron` file:

```{r}
install.packages("rredlist")
usethis::edit_r_environ()   # add the line: IUCN_REDLIST_KEY=your_token
```

Then load it:

```{r setup}
library(aRboretum)
```


## Step 1. Retrieve and review the species data

Start with a character vector containing the species names of interest.

The function `arboretum_data()` queries [Flora e Funga do Brasil (FFB)](https://floradobrasil.jbrj.gov.br/) and the [World Checklist of Vascular Plants (WCVP)](https://powo.science.kew.org/), resolves accepted names and synonyms, adds the [IUCN Red List](https://www.iucnredlist.org/) status of each species, and compiles a structured dataset for downstream use in `aRboretum`.

Both `.csv` and `.xlsx` outputs are supported.

```{r}
species_list <- c("Luetzelburgia bahiensis", "Paubrasilia echinata")

arboretum_data(
  spp_list = species_list,
  printed_lang = c("pt", "en", "fr", "es"),
  save = TRUE,
  format = "xlsx",
  dir = "arboretum_data"
)
```

This saves `arboretum_data/arboretum_data.xlsx` together with `arboretum_data/__data_reviewing_guide.html`. Review the data before going further, either in the spreadsheet or in the guide, which lets you edit the retrieved fields (vernacular names, distribution, endemism in Brazil, establishment means, IUCN status, and so on) and the optional plant uses and notes in your browser, then save or export the corrected file. `printed_lang` sets the languages of the guide.

The FFB dataset is downloaded once into a `flora_download/` folder and reused by later runs, so they also work offline (see the `ffb_dir` argument). Running `arboretum_data()` again on the same `dir` reads the existing data file instead of querying the databases again.

## Step 2. Generate the species phrases

Once you agree with the data, generate the natural-language species descriptions:

```{r}
arboretum_phrases(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en", "fr", "es")
)
```

The phrases are stored in the data file, in the new columns `full_phrases_EN`, `full_phrases_PT`, `full_phrases_ES`, `full_phrases_FR`, and `full_phrases_ADD_LANGUAGE`. The function also writes `arboretum_data/__phrase_generating_guide.html`, where you can read and edit the phrases and export the updated data. Edited phrases are used as they are by the labels and audio guides; running `arboretum_phrases()` again only fills empty phrases (use `overwrite = TRUE` to regenerate them all).

## Optional step. Add personal photos and audio recordings

Personal photos and recordings must be in place before the labels are generated.

```{r}
# One folder per species for photos
arboretum_photos(
  data_path = "arboretum_data/arboretum_data.xlsx"
)

# One folder per species and language for recordings, plus a recording guide
arboretum_audios(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en", "fr", "es")
)
```

Add images to the `arboretum_photos/` folders and recordings to the `arboretum_audios/` folders; the recording guide `arboretum_audios/__personal_audio_recording_guide.html` shows the phrase to read for each species and language. See the [howto_photos](howto_photos.html) and [howto_audios](howto_audios.html) articles for details.

## Step 3. Generate HTML labels

Use the data file, now completed with phrases, to create one interactive HTML label per species. The labels only read the stored phrases and never generate them, so every language in `printed_lang` must have been included in `arboretum_phrases()`.

```{r}
arboretum_labels(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en", "fr", "es"),
  dir = "arboretum_labels"
)
```

Each label can include:

- taxonomic information and authorship;
- species description in multiple languages;
- world and Brazil distribution maps;
- conservation status and source links;
- browser-based text-to-speech;
- personal audio recordings and photos when available.

## Step 4. Build a multilingual minisite

Create a searchable `index.html` page linking all generated species labels.

```{r}
arboretum_minisite(
  labels_dir = "arboretum_labels",
  data_path = "arboretum_data/arboretum_data.xlsx",
  site_title = "My Plant Collection",
  printed_lang = c("pt", "en", "fr", "es"),
  group_by_family = TRUE
)
```

This writes `index.html` into the labels folder and then renames that folder after the site title, here `My_Plant_Collection_minisite/`. Open its `index.html` in your browser to explore the collection homepage. The whole folder can be published as is.

## Optional step. Add one community or local language

In some projects, it is useful to include one additional language in the species labels without translating the full website or minisite interface. This can be especially relevant in collaborative work with Indigenous peoples or other local communities.

For this purpose, `aRboretum` provides the `add_lang` argument in `arboretum_phrases()`, `arboretum_audios()` and `arboretum_labels()`.

To use this workflow:

1. Run `arboretum_phrases()` with `add_lang`: it adds an editable field for the extra language to the phrase guide.
2. Enter the complete label text for each species in the additional language, in the guide or directly in the `full_phrases_ADD_LANGUAGE` column of the data file.
3. Use `add_lang` to include the extra language in the audio-folder workflow and in the final labels.

Example with Tukano:

```{r}
arboretum_phrases(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en"),
  add_lang = "TUKANO"
)

arboretum_audios(
  data_path = "arboretum_data/arboretum_data.xlsx",
  printed_lang = c("pt", "en"),
  add_lang = "TUKANO"
)

arboretum_labels(
  data_path = "arboretum_data/arboretum_data.xlsx",
  audio_dir = "arboretum_audios",
  printed_lang = c("pt", "en"),
  add_lang = "TUKANO",
  dir = "arboretum_labels"
)
```

In this workflow:

- Portuguese and English phrases come from the standard phrase generation;
- Tukano text is read directly from `full_phrases_ADD_LANGUAGE`;
- Tukano does not need to pass through the translation pipeline;
- personal recordings in the `TUKANO` folders are used when available.

See the [add_lang](add_lang.html) article for a complete example.

## What you get

After running the steps above, the minisite folder (`My_Plant_Collection_minisite/` in this example) contains:

1. One HTML file per species, for example `FABACEAE_Paubrasilia_echinata_label.html`.
2. One `index.html` file for the minisite homepage.
3. Supporting folders with maps, icons, and any personal photos and recordings.

Open the species label files in your browser to explore:

- language selection buttons;
- text and audio for each species;
- interactive maps;
- logos and source links;
- personal photos, when provided.

Open `index.html` to explore the collection homepage, including:

- search tools;
- grouped species listings;
- summary cards and dashboard charts built from the species data.

## Next steps

To continue customizing your project:

- add your own photos with the workflow described in the [howto_photos](howto_photos.html) article;
- add personal audio recordings with the workflow described in the [howto_audios](howto_audios.html) article;
- generate QR-code labels for printing with `arboretum_qrcodes()` (see the [aRboretum](aRboretum.html) article).
